---
title: "Thrips survival analysis"
author: "Marc Galland"
date: "`r Sys.Date()`"
output:
    html_document:
        number_sections: yes
        toc: yes
        toc_depth: 2  
params:
  data: "~/SURFdrive/BleekerLab/06.raw_data_from_xp/bioassays/thrips/20161216_survival/survival_data.txt"
  resdir: "results"
  accession2species: "~/SURFdrive/trichome_team/07.data/accession2species.txt"
---

# Background
This script executes a survival analysis ("analysis of time to event") from the thrips survival data (several measurements during consecutive days). The __event__ here is the death of the individual thrips.

This type of analysis is characterized by so-called __censored observations__ which are partial observations.
For instance, some thrips have escaped the observational unit (a plastic box). Thus, we have only partial measurements for these cases. Thrips that survived until the end of the experiment are also censored (we did not observed the event). 

## Online resources
More on this type of analysis can be found:
-  [Statistic courses by German Rodriguez](http://data.princeton.edu/pop509) 
-  [Data blog by steadyfish part 1](http://justanotherdatablog.blogspot.nl/2015/08/survival-analysis-1.html)
-  [Data blog by steadyfish part 2](http://justanotherdatablog.blogspot.nl/2015/08/survival-analysis-2.html)
-  [University of Idaho](http://www.webpages.uidaho.edu/wlf448/2007/LifeTables.htm)
-  [StatSoft only documentation](http://www.statsoft.com/Textbook/Survival-Failure-Time-Analysis#general)

## Kaplan-Meier estimates
From [wikipedia](https://en.wikipedia.org/wiki/Survival_analysis#Kaplan-Meier_plot_for_the_aml_data):
> the Survival function S(t), is the probability that a subject survives longer than time t. 
> S(t) is theoretically a smooth curve, but it is usually estimated using the Kaplan-Meier(KM) curve. 
> The graph shows the KM plot for the aml data.

# Configuration 

## libraries
```{r load libraries,echo=FALSE,warning=FALSE,message=FALSE}
# survminer
if (is.element('survminer', installed.packages()[,1]))
{
  suppressPackageStartupMessages(require('survminer'));
} else
{
  install.packages('survminer');
  suppressPackageStartupMessages(library('survminer'));
}

# survival 
if (is.element('survival', installed.packages()[,1]))
{
  suppressPackageStartupMessages(require('survival'));
} else
{
  install.packages('survival');
  suppressPackageStartupMessages(library('survival'));
}

# dplyr
if (is.element('dplyr', installed.packages()[,1]))
{
  suppressPackageStartupMessages(require('dplyr'));
} else
{
  install.packages('dplyr');
  suppressPackageStartupMessages(library('dplyr'));
}

# svglite
if (is.element('svglite', installed.packages()[,1]))
{
  suppressPackageStartupMessages(require('svglite'));
} else
{
  install.packages('dplyr');
  suppressPackageStartupMessages(library('svglite'));
}
```

## Working directory
```{r working dir}
setwd("/Users/mgalland/SURFdrive/BleekerLab/00.manuscripts/01.natural_insecticides_from_wild_tomatoes/01.Figures/_Figure3_thrips/thrips/")
```

## Data import
```{r data import}
# check file extension to load file
survData = read.delim("~/SURFdrive/BleekerLab/06.raw_data_from_xp/bioassays/thrips/20161216_survival/survival_data.txt",header = T,stringsAsFactors = F)

# change column names
colnames(survData) = c("accession","time","status")

# import accessions to species correspondence
accession2species = read.delim("accession2species.txt",header = T,sep = "\t",stringsAsFactors = F)
```

## Directories 
```{r directories}
# create directories for output (plots...)
resdir = "results"
unlink(resdir,recursive = TRUE)
dir.create(file.path(getwd(),resdir),showWarnings = F)
```


# Survival analysis 

## Medians
Here we want to (i) fit survival curves for each accession. 
Then, we can calculate median survival times `S(t0.5)=0.5` for each accession. 
We order accessions according to this median survival value.
```{r medians}
fit <- with(survData,survfit(formula = Surv(time,status) ~ accession))
# print values
fit
```

```{r extract medians and plot them}
# convert survfit to dataframe
#capture.output(fit,file = file.path(resdir,"survfit.txt"))


# add species
df = read.delim("medians.txt",header=T,stringsAsFactors = F)
df = dplyr::left_join(df,accession2species,by="accession")
df = df[which(!is.na(df$species)),]

# order by biggest median survival time
df = df[order(df$median,decreasing = F),]

# plot
g <- ggplot(df,mapping = aes(x=accession,y=median,fill=species)) +
  geom_bar(stat="identity",colour="black") + 
  scale_x_discrete(limits=df$accession) +
  theme(axis.text.x = element_text(angle=40,hjust = 1)) +
  labs(y = "Median survival time (days)")

print(g)
ggsave("medians.png",plot = g,width = 7,height = 5,dpi=600)
ggsave("medians.svg",plot = g,width= 7,height = 5)


```

## Survival curves
Define a helper function
```{r helper function}
# define helper function to plot survival curve
plotSurvival = function(df,confidence=0.95,xmin=0,xmax = 19){
  # calculate survival object
  survObject = with(df,Surv(time = time,event = status))
  # fit model
  fit <- survfit(formula = survObject ~ accession,data = df,conf.int=confidence)
  # plot model
  g <- ggsurvplot(
   fit,                      # survfit object with calculated statistics.
   risk.table = FALSE,        # show risk table.
   pval = TRUE,              # show p-value of log-rank test.
   conf.int = TRUE,          # show confidence intervals for 
                             # point estimaes of survival curves.
   xlim = c(xmin,xmax),           # present narrower X axis, but not affect
                             # survival estimates.
   break.time.by = 1,        # break X axis in time intervals by 500.
   ggtheme = theme_bw(),  # customize plot and risk table with a theme.
   risk.table.y.text.col = T,  # colour risk table text annotations.
   risk.table.y.text = FALSE, # show bars instead of names in text annotations in legend of risk table
   font.main = 18,              # title font size
   font.x = 16,                 # font x axis 
   font.y = 16                 # font y axis
   )
  return(g)
}
```

Create plots
```{r create plots}
# empty list for plots
l = list()

# make plots
accessions = unique(survData$accession)
for (i in seq_along(accessions)){
  df = dplyr::filter(survData,accession == accessions[i])
  l[[i]] = plotSurvival(df)
  plot_title = accessions[i]
  gg = l[[i]]$plot + ggtitle(plot_title)
  ggsave(filename = file.path(resdir,paste(accessions[i],".png",sep = "")),plot = gg,width = 7,height = 5,dpi = 400)
}
```

## Log rank test
If `rho=0` this is equivalent to log rank test.
If `rho=1` this is equivalent to Peto & Peto test.
```{r log rank test}
survdiff(data = survData,formula = Surv(time,status) ~ accession,rho = 0)
```

## Cox proportional hazards 
Based on the hypothesis that hazard (risk of dying ) depends on the accession.
```{r}
# fit a Cox PH model
fitCox = coxph(formula = Surv(time,status) ~ accession,data = survData)
summary(fitCox)
```

# session info
```{r}
sessionInfo()
```
